---
title: Software install notes
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

Some of The software you will need

## Packages

### Mac Specific
Brew: https://brew.sh/
you will likely need gcc: brew install gcc

### Windows
virtual box: https://www.virtualbox.org/

### Minc Toolkit
bic-mni.github.io

### Pyminc
github.com/Mouse-Imaging-Centre/pyminc

### Minc-Stuffs
github.com/Mouse-Imaging-Centre/minc-stuffs

### RMINC
github.com/Mouse-Imaging-Centre/RMINC

### Pydpiper
github.com/Mouse-Imaging-Centre/pydpiper

## Install Notes

### Pyminc (mac)

```{r, engine = 'bash'}
brew install python3
git clone --recursive https://github.com/Mouse-Imaging-Centre/pyminc
cd pyminc
python3 setup.py install
# you may need to add this to .bash_profile (maybe)
# LD_LIBRARY_PATH=/opt/minc/1.9.15/minc-toolkit-config.sh:$LD_LIBRARY_PATH
```

### RMINC (mac)

Please note that for install RMINC, there is an [INSTALL](https://github.com/Mouse-Imaging-Centre/RMINC/blob/master/INSTALL)
file in the github repo that
has some important notes. Especially for mac users, the Fortran linking notes
are important. The current brew version of gcc as of this morning is
/usr/local/Cellar/gcc/7.2.0/lib/gcc/7 instead of 5.3.0 shown in the INSTALL file.

```{r, engine = 'bash'}
brew install R # only if you don’t have R
```

Do the following all in **the same terminal** (just for installing RMINC)

```{r, engine = 'bash'}
export MINC_PATH=/opt/minc/1.9.15/ # Then run the basic install
R
```

```{r}
install.packages("devtools")
devtools::install_github("Mouse-Imaging-Centre/RMINC", dependencies = TRUE)
```

To test the install run

```{r}
library(RMINC)
install.packages(c("lmerTest", "testthat")) #you probably need these
runRMINCTestbed()
```

If R complains about anatomy hierarchy those errors are spurious, other
errors are potentially problematic.
