---
title: "Basic Stats"
output:
  html_document: default
  html_notebook: default
---

# Load the data

```{r}
library(ggplot2)
load("/hpf/largeprojects/MICe/jason/MISS2017/intro-stats/irData.RData")
```

```{r}
head(irData)
```

This is from the initial radiation study described by Elizabeth this morning, with more information to be found here:

Gazdzinski LM, Cormier K, Lu FG, Lerch JP, Wong CS, Nieman BJ. Radiation-induced alterations in mouse brain development characterized by magnetic resonance imaging. Int J Radiat Oncol Biol Phys. 2012 Dec 1;84(5):e631–8.

Based on that data, we will look at the dentate gyrus of the hippocampus to illustrate some issues in statistical modelling.

# Plotting

First, a few ways of showing the data:

```{r}
qplot(Dose, dentategyrus, data=irData, geom="point") + 
  geom_smooth(method="lm") +
  scale_x_continuous(breaks=unique(irData$Dose)) + 
  ylab(expression(Volume ~ (mm^3))) +
  ggtitle("Volume of Dentate Gyrus", subtitle = "Plotted against Irradiation Dose") + 
  theme_classic() 
```

```{r}
ggplot(irData, aes(y=dentategyrus, x=as.factor(Dose))) + 
  geom_boxplot() + 
  xlab("Radiation Dose") + 
  ggtitle("Volume of Dentate Gyrus") + 
  ylab(expression(Volume ~ (mm^3))) +
  theme_classic()
```


```{r}
library(ggjoy)
ggplot(irData, aes(x=dentategyrus, y=as.factor(Dose))) + 
  ylab("Radiation Dose") +
  xlab(expression(Volume ~ (mm^3))) +
  ggtitle("Volume of Dentate Gyrus") +
  geom_joy()
```

# Modelling

Let's start simple, and reduce it to just a dose of 0 (i.e. sham irradiation) and a dose of 7.

```{r}
irSimple <- subset(irData, Dose %in% c(0,7))
```

And look at what the data shows

```{r}
table(irSimple$Dose, irSimple$Sex)
```


A simple linear model

```{r}
summary(lm(dentategyrus ~ Dose, irSimple))
```

This is in units of dose, i.e. for every Gy you decrease volume by -0.14.

So, highly significant. Let's explore


Significance by simulation:

```{r}
m <- mean(irSimple$dentategyrus[irSimple$Dose==0])
s <- sd(irSimple$dentategyrus[irSimple$Dose==0])

nSims <- 1000
simDraws <- vector(length = nSims)
nSamples <- 10
doseEffect <- -0.13822


for (i in 1:nSims) {
  simDraws[i] <- mean( rnorm(nSamples, mean=m, sd=s))
}
hist(simDraws)
abline(v=m+doseEffect, col="red")
mean(simDraws < (m+doseEffect))
mean(simDraws < (m+(doseEffect*7)))
```

By permutations

```{r}
library(broom)
nSims <- 1000
nSamples <- nrow(irSimple)
doseEffects <- vector(length=nSims)

for (i in 1:nSims) {
  irSimple$doseDraw <- sample(irSimple$Dose, size = nSamples, replace = FALSE)
  l <- lm(dentategyrus ~ doseDraw, irSimple)
  doseEffects[i] <- tidy(l)[2, "statistic"]
}
hist(doseEffects)
mean(doseEffects < -7.917)

```

Let's look at a more subtle effect: sex in mice with sham irradiation:

```{r}
irSex <- subset(irData, Dose == 0)
table(irSex$Sex)
```

```{r}
summary(lm(dentategyrus ~ Sex, irSex))
```

And let's redo the permutation test

```{r}
library(broom)
nSims <- 1000
nSamples <- nrow(irSex)
sexEffects <- vector(length=nSims)

for (i in 1:nSims) {
  irSex$sexDraw <- sample(irSex$Sex, size = nSamples, replace = FALSE)
  l <- lm(dentategyrus ~ sexDraw, irSex)
  sexEffects[i] <- tidy(l)[2, "statistic"]
}
hist(sexEffects)
abline(v=0.03365, col="red")
mean(sexEffects > 0.03365)
mean(sexEffects > 0.03365) * 2
```

# More complex models

Back to the whole data

```{r}
summary(lm(dentategyrus ~ Dose, irData))
```

Let's add sex

```{r}
qplot(Dose, dentategyrus, data=irData, geom="point", colour=Sex) + 
  geom_smooth(method="lm") +
  scale_x_continuous(breaks=unique(irData$Dose)) + 
  ylab(expression(Volume ~ (mm^3))) +
  ggtitle("Volume of Dentate Gyrus", subtitle = "Plotted against Irradiation Dose") + 
  theme_classic()
```


```{r}
summary(lm(dentategyrus ~ Dose * Sex, irData))
```

