---
title: "Intro to stats"
output:
  html_document: default
  html_notebook: default
---

# Load the data and prepare the environment

First, read in the file that contains the description of all the scans in the irradiation study that we will be using as our example dataset.

```{r}
gf <- read.csv("/hpf/largeprojects/MICe/jason/MISS2017/intro-stats/fixed_datatable_IRdose.csv")
```

Take a look at the data in that file:

```{r}
names(gf)
```

The last three columns contain the filenames for the Jacobian determinants, either with or without blurring. Sex and Dose are the most interesting covariates here.

```{r}
table(gf$Dose, gf$Sex)
```

So about 5 males and 5 females for each irradiation dose.

# Exploring the highest dose data

Starting simple, let's just look at a dose of 0 vs 7, for now.

```{r}
d7vsd0 <- subset(gf, Dose %in% c(0, 7))
table(d7vsd0$Dose, d7vsd0$Sex)
```

To do further analyses and see the results, we need to load two more bits of data: the background non-linear average on which we will display results, and the labels for computing volumes.

```{r}
library(RMINC)
anatFile <- "/projects/moush/lbernas/Irradiation_behaviour_project/fixed_build_masked_23mar13_nlin/nlin-3.mnc"

labelFile <-
    paste0("/hpf/largeprojects/MICe/jason/MISS2017/intro-stats/atlas-registration/"
         , "pipeline-18-08-2017-at-07-08-48_processed/nlin-3/voted.mnc")

defsFile <-
    paste0("/hpf/largeprojects/MICe/tools/atlases/"
         , "Dorr_2008_Steadman_2013_Ullmann_2013_Richards_2011_Qiu_2016_Egan_2015_40micron/"
         , "mappings/DSURQE_40micron_R_mapping.csv")
```

# Linear model at every voxel

```{r}
vs <- mincLm(Jacobfile_scaled0.2 ~ Dose, d7vsd0, mask=labelFile)
```

Let's explore the data. First, get a sense of the background range for the anatomical volume, after reading it in, of course.

```{r}
anatVolume <- mincArray(mincGetVolume(anatFile))
hist(anatVolume)
```


```{r, eval = FALSE}
d7vsd0$DoseFactor <- as.factor(d7vsd0$Dose)
launch_shinyRMINC(vs, anatVolume, anatLow=9, anatHigh=15, plotcolumns = d7vsd0)
```


And plot the data with an n of `r nrow(gf)` in my sample.

```{r}
mincPlotSliceSeries(anatVolume, mincArray(vs, "tvalue-Dose"), anatLow=9, anatHigh=15, low=2, high=7, symmetric=T, begin=50, end=-40, legend="t-statistic", plottitle = "Dose (7 vs 0 only)")
```

Now, use the False Discovery Rate to correct for multiple comparisons

```{r}
qvs <- mincFDR(vs, mask=labelFile)
```

Let's see the thresholds:

```{r}
qvs
```

And plot at that threshold

```{r}
mincPlotSliceSeries(anatVolume, mincArray(vs, "tvalue-Dose"), anatLow=9, anatHigh=15, low=3.42, high=7, symmetric=T, begin=50, end=-40, legend="t-statistic", plottitle = "Dose (7 vs 0 only)")
```

Or plot the q values directly

```{r}
mincPlotSliceSeries(anatVolume, -log10(mincArray(qvs, "tvalue-Dose")), anatLow=9, anatHigh=15, low=1.3, high=3, symmetric=F, begin=50, end=-40, legend="-log10 q value", plottitle = "Dose (7 vs 0 only)")
```

Or the same, but with direction added again:

```{r}
mincPlotSliceSeries(anatVolume, -log10(mincArray(qvs, "tvalue-Dose"))*sign(mincArray(vs, "tvalue-Dose")), anatLow=9, anatHigh=15, low=1.3, high=3, symmetric=T, begin=50, end=-40, legend="log10 q value", plottitle = "Dose (7 vs 0 only)")
```

# Linear model across volumes

Now compute the volumes. There are several ways of doing so; in this case, we'll integrate the Jacobian determinants, since that is what we have for this particular pipeline. Takes a few minutes to run.

```{r}
library(RMINC)
allvols <- anatGetAll(d7vsd0$Jacobfile_scaled, atlas=labelFile,
                      defs=defsFile,method="jacobians")
```

And we have to put anatomy and hierarchical definitions together. This takes a few minutes to run

```{r}
AllenDefs <- "/hpf/largeprojects/MICe/tools/atlases/Allen_Brain/Allen_hierarchy_definitions.json"
hdefs <- makeMICeDefsHierachical(defsFile, AllenDefs)
```

Next, add the volumes from this dataset into the hierarchy.

```{r}
hanatd7vsd0 <- addVolumesToHierarchy(hdefs, allvols)
```

Run a simple linear model

```{r}
hVs <- hanatLm(~Dose, d7vsd0, hanatd7vsd0)
hanatView(hVs, "tvalue.Dose", low=2, high=6, symmetric=T)

```

Print them to the screen


```{r}
print(hVs, "tvalue.Dose", limit=NULL)
```

And display the results as images.

```{r}

labelVolume <- mincArray(mincGetVolume(labelFile))

vol <- hanatToVolume(hVs, labelVolume, "tvalue.Dose")
mincPlotSliceSeries(anatVolume, vol, low=2, high=6, symmetric=T, anatLow=9, anatHigh=15)
```



Now, let's correct for multiple comparisons here too:

```{r}
hVs <- hanatFDR(hVs)
thresholds(hVs)
```

```{r}
hanatView(hVs, "tvalue.Dose", low=4.12, high=8, symmetric=T, fontsize = 40, levelSeparation = 1000)
```

```{r}
vol <- hanatToVolume(hVs, labelVolume, "tvalue.Dose")
mincPlotSliceSeries(anatVolume, vol, low=4.12, high=8, symmetric=T, anatLow=9, anatHigh=15)
```



Let's do the anatomical analysis only on merged left and right variants of each structure.

```{r}
library(data.tree)
hVsSymmetric <- Clone(hVs)
Prune(hVsSymmetric, function(x) !(startsWith(x$name, "right") | startsWith(x$name, "left")))
print(hVsSymmetric, "tvalue.Dose", limit=NULL)
                   
```

```{r}
vol <- hanatToVolume(hVsSymmetric, labelVolume, "tvalue.Dose")
mincPlotSliceSeries(anatVolume, vol, low=4.12, high=8, symmetric=T, anatLow=9, anatHigh=15, 
                    begin=50, end=-50)
```

We can also look at just a subset of the tree:

```{r}
print(FindNode(hVsSymmetric, "Hippocampal region"), "tvalue.Dose", limit=NULL)
```

```{r}
vol <- hanatToVolume(FindNode(hVsSymmetric, "Hippocampal region"), labelVolume, "tvalue.Dose")
mincPlotSliceSeries(anatVolume, vol, low=4.12, high=8, symmetric=T, anatLow=9, anatHigh=15, 
                    begin=120, end=-170, plottitle = "Dose, Hippocampal region", legend="t-statistic",
                    mfrow = c(3,3))
```


```{r}
library(ggplot2)
d7vsd0$volume <- FindNode(hanatd7vsd0, "Dentate gyrus")$volumes
qplot(as.factor(Dose), volume, data=d7vsd0, geom="boxplot")
```

```{r}
library(RMINC)
allvols <- anatGetAll(gf$Jacobfile_scaled, atlas=labelFile,
                      defs=defsFile,method="jacobians")
hanat <- addVolumesToHierarchy(hdefs, allvols)
```

