---
tile: Intro to bash
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(engine = 'bash')
```

## What is a command line or terminal?

- Great overview and tutorial: [](bash_tutorial.pdf)
- Open terminal 
	- on Mac: command + space -> Terminal
	- on Windows: use ssh client, e.g. cygwin 
- Basically a text-based interface
- Can enter commands & feedback given as text

### why bother?

- manipulate data
- run registration pipeline in terminal
- faster to navigate through folders

## ssh

- you can use terminal to log into other computers connected to same network
- ssh: secure shell
- login to network here using following username + password

```{r, eval = FALSE}
dvousden  password: dvousden123
ssh username@location
ssh -X username@location
```

- first make sure everyone connected to wifi (Chris)
- now try logging into our network

```{r, eval = FALSE}
ssh username@172.16.128.140
```


## The Shell, Bash
- within a terminal you have what is known as shell: defines how terminal will behave & runs commands
- most common is bash
- can check type of shell you have by looking at environment variable.
- use echo command

```{r}
echo $SHELL
```

- make sure says ‘bash’ 

## Shortcuts

- commands stored in history
- push up arrow to go backwards through history
- `history` 
- ctrl + R to search history 

## Moving around the system 

```{r} 
pwd  # tells you where you are, same structure as folders that you’re used to clicking
ls  # shows you what’s there 
```

## getting help &  giving commands arguments

- a lot of commands also take arguments
- commands also have help pages or manuals to show you how to use them
- `man` 
- q to quit

```
ls [-ABCFG…] [file ..]

- tells you that ls ‘list directory contents’
- possible arguments are all the letters
- you can also give ls the name of a particular file and it will give you information about that file
```

```{r} 
ls -l  # gives you long output (who created file, date of creation, size of file) 
#ls -l DIRECTORY # shows you what’s in that directory 
```

## Paths

- whenever we refer to a file or directory on a command line, we have to use that file’s path
- path is like the address, where to find the file or folder
- absolute vs. relative paths
- directories are organized hierarchically
	- top is root /
	- absolute paths specify location starting from root

```{r, eval = FALSE}
ls /home/Dulcie/Documents 
```

- relative paths specify location in relation to where you currently are

```{r, eval = FALSE}
#ls Documents 
ls ~   # shortcut for home directory
ls .    # shortcut for current directory
ls ../  # shortcut for parent directory 
```
  
## Moving around

```{r}
pwd
cd ..
pwd 
```
- shortcut: tab completion 

```{r}
# cd Doc   # show tab completion
```

## Create new directory 

```{r, eval = FALSE}
mkdir banana
cd banana 
pwd
ls
cd ..
rmdir banana
```

## Environment variables

- There are a number of different environment variables
- one important one is called PATH 

```{r} 
echo $PATH
```

- this tells the computer where to look for programs and software packages
- if you are having problems running a certain program, sometimes it’s because you haven’t
told the computer where to look

- environment variables can be specified in .bashrc 

```{r, eval = FALSE} 
less ~/.bashrc
```

## The MICe treasure hunt

```{r}
mkdir /tmp/treasure
cd /tmp/treasure
perl /axiom2/projects/software/arch/linux-3_2_0-36-generic-x86_64-eglibc-2_15/bin/treasureHunt.pl
```


Let's have a look at the first clue

```{r}
ls 
```

when done you can delete directory 

```{r}
rm -rf /tmp/treasure
```

## Troubleshooting login

Offending RSA key in /Users/Dulcie/.ssh/known_hosts:12

```
 sed -i '6d' ~/.ssh/known_hosts
 perl -pi -e 's/\Q$_// if ($. == 6);' ~/.ssh/known_hosts
```
